---
title: "Briefing paper on Ethnicity and COVID-19 status in Scotland"

author: Modelling & Research Cell, Health Protection Scotland 

date: "`r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"

header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}

output:
  pdf_document

# rmarkdown::render("Covid_ethnicity_Scotland.Rmd")

---

# Background
There have been anecdotal reports of an excess of hospitalisations, intensive care unit need and death from COVID-19 among non-white ethnic groups in particular Black or Black British and South Asian.  A preliminary study using UK Biobank data has [reported](http://medrxiv.org/cgi/content/short/2020.04.22.20075663) that Black, South Asian and White Irish people were more likely to have confirmed infection (RR 4.01 (95%CI 2.92-5.12); RR 2.11 (95%CI 1.43-3.10); and RR 1.60 (95% CI 1.08-2.38) respectively) 

Understanding whether such inequity in risk of COVID-19 by ethnicity pertains in Scotland and whether disease severity is worse is relevant to early anticipation of health care needs, triaging algorithms being used by NHS Hubs, and shielding policies.

# Methods 
All persons to date  who tested positive  for Sars-CoV-2 RNA were identified using the ECOSS virology laboratory systems. This system records CHI but not ethnicity.

A matched set of 10 controls per case matched for age, sex and GP practice was identified using the CHI database as part of the REACT-COVID19 case control study led by the Modelling & Research cell. 

## Coding of ethnicity
Names of cases and controls were available to the CHI team. They used the ONOMAP software to assign predicted ethnicity to all cases and controls. This was output in anonymised format to the Modelling & Research Cell linkage team. ONOMAP is a software package developed by Mateos at UCL which used people's first and family names to estimate underlying ethnic group. It has been developed using algorithms built from the Worldnames database that covered both names and ethnic group in a large number of countries. ONOMAP assigns both predicted geographical region of origin and ethnicity. ONOMAP has previously been reported to have variables sensitivity for assigning non white ethnic groups 17-89%. ([Lakha 2011](https://www.ncbi.nlm.nih.gov/pubmed/21907365)) The ONOMAP "Onolytic Types" were allocated to ethnic groups. Names classified "Muslim" in the ONOMAP classification were assigned to the South Asian group, as Scotland most people with Muslim names are South Asian.  The ONOMAP classification assigns a high proportion of names in Scotland to the category "Celtic" which does not distinguish Irish from Scottish names so that White Irish cannot be differentiated from White British.

Hospital discharge and outpatient attendance data (SMR01 and SMR00 respectively) have increased in their completion of mandatory ethnic group in recent years. For all cases and controls the linkage team extracted i) the most recent ii) the most frequent ethnic group assignation from these two datasets  with a 10 year lookback period. Unknown or missing codes were excluded.  Concordance between the most recent SMR ethnic group   and the most frequent  SMR ethnic group was 98%. 

A previous comparison of ONOMAP against SMR01 records in Scotland concluded that SMR01 records with a missing ethnic group code were broadly representative of records with ethnicity recorded.  (Clark & Buchanan ISD 2017)
The classification given below follows the Census categories, collapsed into five groups: White, Black, South Asian, Chinese, Other. 

# Results

## Agreement between ethnicity assigned by name classification (ONOMAP) and self-reported ethnicity in  SMR records

```{r ethnic, echo=FALSE, warning=FALSE, message=FALSE}

pander(table.ethnic, missing=".",
       split.tables=Inf,
       split.cells=rep(16, 6), 
       caption="\\label{tab:crosstab}Comparison of ethnicity assigned by ONOMAP name classification (rows) with ethnicity assigned in SMR  records (columns)")

```

**Amanda see line below for example of how to embed numbers from R in text**

Table \ref{tab:crosstab} shows that the ONOMAP name classification assigns to the same as self-reported category `r testpositives.ethnic[1, 1]` 


of those who self-report their ethnicity as White, 92% of those who self-report as South Asian, 71% of those who self-report as Chinese, but only 21% of those who self-report as Black.  

In all further analyses, the ONOMAP classification has been collapsed to four groups: White, South Asian, Chinese, Other.  Analyses that distinguish the Black group can be done only for the subset of individuals who have SMR records. 


## Relation of severity of COVID-19 to ethnicity among test-positive cases


```{r testpositives.ethnic, echo=FALSE, warning=FALSE, message=FALSE}

pander(testpositives.ethnic, missing=".",
       split.tables=Inf,
       split.cells=rep(25, 3), 
       caption="Severity of disease in test-positive cases, by ethnicity assigned from ONOMAP name classification")

```

```{r testpositives.ethnic.smr, echo=FALSE, warning=FALSE, message=FALSE}

pander(testpositives.ethnic.smr, missing=".",
       split.tables=Inf,
       split.cells=rep(25, 3), 
       caption="Severity of disease in test-positive cases, by ethnicity assigned in SMR records")

```

## Relation of severe / fatal COVID-19 to ethnicity in the Scottish population

```{r onomap, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.demog.aug) <- gsub("_", " ", rownames(table.demog.aug))
knitr::kable(table.demog.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="onomap",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value",
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with demographic variables including ethnicity based on name classification") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    landscape()
```

```{r ethnicsmr, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.ethnicsmr.aug) <- gsub("_", " ", rownames(table.ethnicsmr.aug))
knitr::kable(table.ethnicsmr.aug,
             escape=FALSE, 
             booktabs=TRUE,
             label="ethnicsmr",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.ethnicsmr.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "p-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with ethnicity in those with ethnicity recorded in Scottish Morbidity record") %>%
    column_spec(column=1, width="4.5cm")
```


As shown above the numbers of persons testing positive from non-White ethnic groups is low. There is no evidence that the frequency of severe disease among those who test positive is any higher in non-White groups than in Whites. The frequency of severe disease was slightly lower in South Asans but with these small numbers this estimate is unstable. 

# Conclusion
Although the numbers are small, there is no evidence that the main ethnic minorities in Scotland are at higher risk of severe COVID-19 than the  White  population. The point estimates for severe disease are lower in the main minority ethnic groups than in Whites.  However power to detect anything other than very large differences is limited as yet given the low numbers of non-White persons tested positive.  Therefore we will continue to monitor the situation. 

Recording of ethnicity in hospital admission does not provide complete coverage of the population at risk of COVID-19.  The coding is also incomplete with substantial missingness.  Merging ethnicity data from the Scottish Morbidity Record with name based classification would introduce differential misclassification of ethnicity by health status.  ONOMAP/ name based classification does not reliably distinguish people of Black-Caribbean heritage from people of European ancestry.  Access to primary care data would likely increase the completeness with which ethnicity is assigned but governance issues need resolution.

A possible explanation for reports from England that people from minority ethnic groups were over-represented among cases of severe and fatal disease may be that these groups make up a high proportion of front-line health care workers and care givers in England, who were heavily exposed to infectious cases.  In Scotland a "healthy migrant effect", such that those who migrate tend to be fitter than those who do not migrate, may explain the low rates of severe disease in minority groups.  
