---
title: "Risk factors for severe COVID-19 in the Scottish population: the REACT-SCOT Study"
author: Paul McKeigue 
date: "`r gsub('^0', '', format(Sys.Date(), '%d %B %Y'))`"
output: 
  beamer_presentation:
    slide_level: 1
    includes:
      in_header: ../beamer_header.tex
    latex_engine: lualatex
    keep_tex: yes
---

# REACT-SCOT Study

* Case-control study of all diagnosed cases of COVID-19 in Scotland since March 2020

  - Cases include those without a positive test, ascertained through hospital discharge diagnosis or death certificates. 
  
  - Ten controls per case, not previously diagnosed and matched for age, sex and general practice

  - linkage to household composition, medical records including hospital admissions, prescribing in primary care

* Main outcome variable: **severe COVID-19** defined as entry to critical care within 21 days of presentation date, or fatal outcome

  - definition ensures that case ascertainment is not affected by triage of those assessed as unlikely to benefit from critical care  

# Studies based on REACT-SCOT

* Co-morbidity and construction of a risk score ("COVID age")

* Pharmacoepidemiology: medications compromising COVID

* Effectiveness of shielding clinically extremely vulnerable individuals

* Hospital transmission

* Other studies published, under review or in progress: 
  - associations with occupation: patient-facing health care workers, teachers (inverse association), 
  - (inverse) association with living with children
  - vaccine efficacy in clinically vulnerable individuals 

# Laporte & Healy (April 2020): "Medications compromising COVID"

Drug classes associated with community-acquired pneumonia

* proton pump inhibitors

* hypnotics, sedatives 

* antipsychotics, drugs used in nausea and vertigo 

* opioid analgesics 

* gabapentinoids 

* non-steroidal anti-inflammatory drugs.

* anticholinergic effects likely to increase risk of pneumonia: 

  - H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used for parkinsonism, and antiepileptic drugs
  
Predictions of associations with severe COVID-19 were confirmed in REACT-SCOT.  

# Conditions designated as eligible for shielding (~4% of population)

1. Solid organ transplant recipient

2. Cancer of blood or bone marrow, or cancer treatments that affect immune system 

3. Severe respiratory conditions: cystic fibrosis, asthma, chronic obstructive airway disease, bronchiectasis, pulmonary hypertension

4. Rare diseases that increase risk of infections: severe combined immunodeficiency, sickle-cell disease. 

5. Immunosuppression therapies sufficient to increase risk of infection

7. Other conditions including dialysis, splenectomy

First batch of shielding letters sent on 4 April 2020 advised strict isolation, offered support with home delivery of food and medication. 
 
# Severe cases by risk category and date of presentation: ribbon plot
```{r plotcasesbydate, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.5,fig.asp=0.6, fig.cap="Care home residents excluded"}

options(knitr.kable.NA = '')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

theme_set(theme_gray(base_size = 10))

p.byelig + theme(legend.position = c(0.4, 0.7)) + theme(legend.text=element_text(size=8)) 

```

# Rate ratios for severe COVID-19 associated with shielding and moderate risk condition 
```{r plotrateratio, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.8,fig.asp=0.58, fig.cap="Care home residents excluded. Line thickness is proportional to precision of estimate. Arrows indicate dates that shielding advice letters were sent."}

theme_set(theme_gray(base_size = 10))

p.rateratio + theme(plot.title = element_blank())  + 
labs(caption = element_blank(), tag = element_blank()) +   
theme(legend.position = c(0.48, 0.87)) + theme(legend.text=element_text(size=8)) + 
theme(legend.title = element_blank())

```

# Risk factors for severe COVID-19 in those eligible for shielding 
\scriptsize
\setbeamertemplate{caption}{\raggedright\insertcaption\par}

```{r riskfactorsshieldedonly, echo=FALSE, warning=FALSE, message=FALSE}

library(knitr)
library(kableExtra)

options(knitr.table.format = "latex") # adds tab: prefix to labels

knitr::kable(table.shieldedonly.severecases.nocare[7:15, c(1:2, 5:6)], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="coxfatal",
             row.names=TRUE,
             align=c("r", "r", "l", "l"), 
   col.names=c(clean.header(colnames(table.shieldedonly.severecases.nocare)[1:2]),
                "Rate ratio (95\\% CI)", "p-value")) %>%
    kableExtra::group_rows("School-age children in household", 2, 4, bold=FALSE) %>%
    kableExtra::group_rows("Adults in household", 5, 7, bold=FALSE) %>%
    column_spec(column=1, width="2.3cm") %>%
    column_spec(column=2, width="1.7cm") %>%
    column_spec(column=3, width="1.5cm") %>%
    column_spec(column=4, width="2.2cm") %>%
    column_spec(column=5, width="1.3cm")
 
```

# Prevalence of recent hospital exposure by risk category 

```{r plotexphosp, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.25, fig.asp=0.7, fig.cap="Prevalence in controls not resident in care homes"}
p.hosp + theme(plot.title = element_blank()) + 
labs(caption = element_blank(), tag = element_blank()) + 
theme(legend.position = c(0.4, 0.82))  

```

# Fraction of severe COVID-19 attributable to recent hospital exposure 

```{r plotfrac, echo=FALSE, warning=FALSE, message=FALSE, fig.width=4.25, fig.asp=0.7, fig.cap="Care home residents excluded"}
p.fracattr + theme(plot.title = element_blank())+ 
labs(caption = element_blank(), tag = element_blank())

```

# Conclusions

* Protection of vulnerable individuals was limited by their inability to avoid transmission in hospital and from other adults in the household. 

  - Efficacy of single dose vaccine ~ 70% in shielded but a larger study is required to estimate efficacy in solid organ transplant recipients. 

* Hospital-acquired infection accounted for more than half of severe cases in early December 2020. 

* Role of nosocomial transmission was recognized early in the epidemic: 

  - Wang (2020 Feb) JAMA: 41% of hospitalized cases in Wuhan in January 2020
  - Boccia (2020 Apr) JAMA Int Med: "largely a nosocomial infection" in Lombardy 

* Lessons to be learned: physically separate hospitals, surveillance, epidemic modelling. 




