---
title: 'Associations of severe COVID-19 with drug prescribing in the REACT-SCOT case-control study'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Janet Murray
    affiliation: HPS 
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health. 
geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ../covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::plos_article:
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
#    toc: true
#    toc_depth: 2
csl: ../plos.csl
urlcolor: red

# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf")

---

```{r functions, echo=FALSE}

source("helperfunctions.R")

knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

library(knitr)
library(kableExtra)

```

# Abstract

_Background_-- We previously reported from a case-control study that encashment of any prescription from primary care in the past year was strongly associated with severe COVID-19, independently of the presence of conditions designated as high risk.  The objective of this study was to distinguish between three possible explanations for this unexpected association: confounding by other high-risk conditions, an effect associated with consultations for symptoms not necessarily associated with high-risk conditions, and a direct effect of exposure to one or more classes of drug on risk of severe disease.  

_Methods_ -- Severe cases were ascertained as all those who tested positive for SARS-C0V-2 viral RNA in the national database who entered intensive care or died within 28 days.  We selected 10 controls per case from the national population register, matched for sex, age and general practice.  For all cases and controls we extracted International Classification of Disease (ICD-10) codes from the previous five years of hospital discharge records and British National Formulary (BNF) drug codes of encashed primary care issued prescriptions in the past year.  Rate ratios were estimated by conditional logistic regression.

_Findings_ -- The rate ratio associated with encashment of at least one prescription in the last year was elevated `r table.scrip.any.carehome[1, 3]` in those not resident in a care home but not in care home residents `r table.scrip.any.carehome[2, 3]`, of whom `r table.scrip.any.carehome[2, 1]` of controls had encashed a prescription in the last year. On restriction to those not resident in a care home, the rate ratio associated with encashment of a prescription was elevated in those without any of the designated high-risk conditions `r table.scrip.any.nocare.listed[1, 3]`, but not in those with any designated condition `r table.scrip.any.nocare.listed[2, 3]` of whom `r table.scrip.any.nocare.listed[2, 1]` of controls had encashed a prescription in the last year. 

Among those without any of the designated high-risk conditions (diabetes, heart disease, obstructive airway disease, chronic liver disease, chronic renal failure, neurological including dementia, immune deficiency / suppression) severe cases were highly significantly more likely than controls to have had prescriptions in the past year from each BNF chapter. The strongest associations (rate ratio, 95% CI) were seen with chapters for drugs acting on the gastro-intestinal system `r table.drugs.nocare.notlisted[1, 3]`, cardiovascular system `r table.drugs.nocare.notlisted[2, 3]`, central nervous system `r table.drugs.nocare.notlisted[4, 3]` and "Nutrition and Blood" `r table.drugs.nocare.notlisted[9, 3]`.  When each of these were further disaggregated to subchapter or paragraph level, key findings included a significantly greater risk of severe COVID-19 associated with proton pump inhibitor use with rate ratios varying from `r table.protonpump[4, 3]` in those aged 75 years and over to `r table.protonpump[1, 3]` in those aged under 40 years.  However many other drug classes associated with prescribing for non-specific symptoms were also associated with severe COVID-19: these included osmotic laxatives non-opioid analgesics, oral iron supplements and antipsychotic drugs. 

Among drugs prescribed for cardiovascular disease, only loop diuretics and calcium channel blockers were independently associated with risk of severe COVID-19 in a multivariate analysis. Associations with ACE inhibitors and angiotensin-II receptor blockers were similar to those for other antihypertensive agents. 

*Conclusions* -- Risk of severe COVID-19 is strongly associated with encashment of prescriptions for drugs that are widely prescribed for minor symptoms.  It remains to be established whether this effect is accounted for by association with characteristics of patients who consult for these symptoms, or by direct drug effects.  Of the drug classes that might possibly account for an association with prescribing, by far the strongest candidate is proton pump inhibitors, which are prescribed to about 30% of those aged over 60 years and are associated with a threefold elevation of risk of severe disease in this age range.  As increased risk of infection is an established adverse effect of these drugs, a causal effect is plausible.  This may be relevant to high rates of severe disease reported in other populations where this class of drugs is widely overprescribed.  Until further studies are available, it is advisable to avoid using these drugs without a recognized medical indication.  

\newpage

# Background
A striking preliminary finding of the REACT-SCOT study was the very strong association of severe COVID-19 with having encashed at least one prescription in the past year. The rate ratio associated with this variable varied from `r tables.agegr[[1]][2, 3]` in those aged under 60 years to `r tables.agegr[[3]][2, 3]` in those aged 75 years and over.  In a multivariate analysis adjusting for care home residence, any hospital admission and listed conditions, these rate ratios were reduced to `r tables.agegr[[1]][2, 5]` and `r tables.agegr[[3]][2, 5]` About one third of those over 75 had not encashed a prescription in the previous year.  The objective of this study was to distinguish between three possible explanations for this unexpected association: confounding by other high-risk conditions, an effect associated with consultations for symptoms not necessarily associated with high-risk conditions, and a direct effect of exposure to one or more classes of drug on risk of severe disease.  

# Methods
The design of the REACT-SCOT case-control study is described in detail in the accompanying paper. 

## Case ascertainment and sampling of controls 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day and the weekly updated Scottish Morbidity Record 01. Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care units across Scotland. Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.   

For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case using the Community Health Index (CHI) database. As this is an incidence density sampling design, it is possible and correct for an individual to appear in the dataset more than once, initially as a control and subsequently as a case.  With this design, odds ratios in conditional logistic regression models are equivalent to rate ratios.  

## Morbidity and drug prescribing
For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-Cov-2 and using all codes on the discharge. Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.  

British National Formulary (BNF) drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 25 days before testing positive for SARS-CoV-2.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits) and subparagraph (1 digit), and chemical substance (2 characters).  For this analysis prescription codes from chapters 14 and above, comprising vaccines, anaesthesia, other preparations, dressings and appliances, were grouped as "Other" 

# Data analysis 
We derived indicator variables for a list of conditions that have been included as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material. 
Rate ratios were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.  

## Information governance
This study was conducted under approvals from the Public Benefit and Privacy Panel that allow Health Protection Scotland staff to link any PHI controlled datasets <!--- janet to add detail and reference numbers ---> and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.  

# Results
## Encashment of any prescription by care home status and presence of any listed condition
As residence in a care home was the strongest risk factor for severe COVID-19, we first examined the association with enchashment of any prescription after stratifying by care home status.  Table \ref{tab:carehome} shows that the association was restricted to those not resident in a care home. 

As a next step, we examined the association with encashment of any prescription after stratifying by the presence or absence of a listed condition, excluding those resident in a care home.  Table \ref{tab:listed.any} shows that the association was restricted to those without a listed condition 

## Partitioning by BNF chapter
We next partitioned the "any prescription" variable into indicator variables for each chapter of the British National Formulary.  The chapter level groups drugs by the system on which they act.  We restricted this analysis to those without any listed condition and not resident in a care home.  Table \ref{tab:drugs} shows these associations.  

In univariate analyses, prescriptions in almost all BNF chapters were associated with severe disease. 

Table \ref{tab:explained.chapter} examines how much the effect of any prescription is reduced by adding an indicator variable for each chapter, one at a time.  The chapter indicators that explained the highest fractions of the "any prescription" effect were Chapter 1 (Gastrointestinal System) , Chapter 2 (Cardiovascular System), Chapter 4 (Central Nervous System), and Chapter 9 (Nutrition), The codes grouped under Chapter 14+ (Other) are mostly for dressings and appliances, presumably reflecting a general association with frailty.  This led us to examine associations with specific drug classes, grouped at BNF paragraph or subparagraph level, within each of these three chapters.

## Associations with severe COVID-19 at section, paragraph and subparagraph level

## Drugs for gastro-intestinal system: BNF chapter 1
Table \ref{tab:bnfchapter1} shows that encashed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19.  Of these by far the most commonly prescribed drug classes was proton pump inhibitors.  Osmotic laxatives were also associated with severe disease, but were far less commonly prescribed. 

## Drugs for cardiovascular system: BNF chapter 2
Table \ref{tab:bnfchapter2} lists associations with drugs for the cardiovascular system, grouped by BNF subparagraph.  This shows that prescriptions of most classes of drugs for the cardiovascular system were associated with severe COVID-19. In a multivariate analysis the strongest associations were with loop diuretics and calcium-channel blockers.  All commonly-used drug classes used to treat hypertension were associated with moderately elevated risk of severed disease, with rate ratios from ranging from about 1.4 to 2.2.  The rate ratios associated with ACE inhibitors and angiotensin-II receptor blockers were in the middle of this range. Prescriptions of anticoagulant, antiplatelet and lipid-lowering drugs were associated with severe disease in univariate analyses, but these association were weaker in a multivariate analysis.  

## Drugs acting on the central nervous system: BNF chapter 4
Table \ref{tab:bnfchapter4} shows that encashed prescriptions of most classes of drugs acting on the central nervous system were associated with severe COVID-19. The most commonly prescribed drug class, and the strongest independent effect in a multivariate analysis, was non-opiate analgesics and compound preparations. This class of drugs includes paracetamol and various compound preparations that are marketed for relief of upper respiratory symptoms, headache, influenza, and other conditions.  Other drug classes associated with severe COVID-19 in a multivariate analysis were antipsychotic agents and opioid analgesics.  

## Drugs for Nutrition and Blood: BNF chapter 9
Table \ref{tab:bnfchapter9} shows that prescriptions of several classes of nutritional supplements were associated with severe disease: these included oral iron supplements, drugs used in megaloblastic anemias, enteral nutrition products, and Vitamin D.  

## Association with proton pump inhibitors
As drugs for the gastrointestinal system accounted for the largest fraction of the effect of "any prescription", and proton pump inhibitors were the most commonly prescribe drug in this class, we examined the association with proton pump inhibitors more closely.  Table \ref{tab:protonpump.adjusted} shows the association with proton pump inhibitors by age group, without and with adjustment for NSAIDs, antiplatelet drugs and any past hospital diagnosis in the ICD subchapter "Disorders of esophagus, stomach and duodenum".  The association was strongest in those aged under 40 years, and was reduced by adjusting for covariates, especially prescription of nonopiate analgesics.  

Table \ref{tab:analgesic.adjusted) shows the univariate and multivariate associations of severe COVID-19 with prescription of nonopioid analgesics .  



# Discussion 
## Association with prescriptions for symptoms

The list of prescribed drug classes that are associated with elevated risk of severe COVID-19 correspond closely to those identified by clinical pharmacologists and pharmacists as widely overprescribed drug classes: these include proton pump inhibitors, laxatives, analgesics, antipsychotic agents and nutritional supplements. 

Patients who consult with minor symptoms, leading to prescriptions of these drugs without a clear medical indication, are at high risk.  The problem is to determine whether this association is accounted for by characteristics that underlie these consultations for minor symptoms, or by direct drug effects.  

Because this case-control study is matched on primary care practice, variation of over-prescribing between practices does not contribute to the conditional odds ratios. 

## Proton pump inhibitors
The association with proton pump inhibitors is not explained by association any underlying gastrointestinal disease that leads to hospital admission, or by coprescribing with non-steroidal antiinflammatory or antiplatelet agents.  As proton pump inhibitors are available over the counter, our study of encashed prescriptions does not capture all exposure to this class of drugs.  However as there are no prescription charges in Scotland, long-term users of these expensive drugs are likely to seek to obtain them on prescription.  

As proton pump inhibitors are associated with increased risk of enteric infections, it is plausible that they could increase the risk of severe infection with SARS-CoV-2, which is at least partly an enteric infection [@wan_enteric_2020].  Others have noted that proton pump inhibitors are widely prescribed among the elderly, often for no clear indication [@marks_time_2016; @kurdi_ongoing_2018].  35% of patients admitted to acute geriatric units in Aberdeen had been inappropriately prescribed proton pump inhibitors: overprescribing was associated with the number of other drugs prescribed and with an index of comorbidity [@jarchow-macdonald_prescribing_2013]. 

In Szechuan, China, PPIs were prescribed to 37% of inpatients in a tertiary referral centre in 2016 [@luo_changes_2018]. A Spanish drug-utilization study showed that 55% of hospital inpatients were presribed a PPI at discharge [@ramirez_overuse_2010].  

## Cardiovascular drugs
There is evidence of any effect of ACE inhibitors or angiotensin-II receptor blockers beyond the associated with treated hypertension.  

## Methodological strengths and weaknesses

We have prescription encashments, not just prescriptions.  Drug coding by BNF codes is more systematic than SNOMED

We don't have all the morbidity data available in the primary care record but we have hospital diagnoses which are ICD-coded by trained coders.  


## Conclusion
It is not possible to establish unequivocally whether the association with prescription encashment is accounted for by patient characteristics associated with prescribing for minor symptoms, or by a direct effect of any specific class of drugs.  However of the drug classes that are associated with increased risk that is not explained by underlying high-risk conditions, proton pump inhibitors are the strongest candidate for a causal effect.  Until further studies are available, it is advisable to avoid using drugs in this class without a recognized medical indication.  

# References {-}

<div id="refs"></div>

```{r carehome, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.scrip.any.carehome,
             escape=FALSE, 
             booktabs=TRUE,
             label="carehome",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.scrip.any.carehome)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with encashment of any prescription in the last year, by care home status") %>%
     kable_styling(latex_options="hold_position") %>%
   column_spec(column=1, width="4cm")
```

```{r listed, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.scrip.any.nocare.listed,
             escape=FALSE, 
             booktabs=TRUE,
             label="listed.any",
             col.names=c(clean.header(colnames(table.scrip.any.nocare.listed)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with encashment of any prescription in the last year, by presence of a listed condition, excluding care home residents") %>%
    kable_styling(latex_options="hold_position") %>%
    column_spec(column=1, width="4cm")
```

```{r drugs, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.drugs.nocare.notlisted) <- gsub("_", " ", rownames(table.drugs.nocare.notlisted))
knitr::kable(table.drugs.nocare.notlisted,
             escape=FALSE, 
             booktabs=TRUE,
             label="drugs",
             digits=c(0, 0, 0, 200, 0, 200), 
             col.names=c(clean.header(colnames(table.drugs.nocare.notlisted)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in those not resident in care homes and without listed conditions") %>%
    column_spec(column=1, width="4.5cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```


```{r explained.chapter, echo=FALSE, warning=FALSE, message=FALSE}

rownames(table.anyscrip.chapter) <- gsub("\\_", " ", rownames(table.anyscrip.chapter))
knitr::kable(table.anyscrip.chapter,
             escape=FALSE, 
             booktabs=TRUE,
             label="explained.chapter",
             col.names=c("Coefficient", "Rate.ratio", "Proportion explained by covariate"), 
             digits=c(2, 2, 2), 
             caption="Effect of an indicator variable for each BNF chapter on the size of effect associated with encashment of any prescription, in those not in care homes and without listed conditions") %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="2cm") %>%
    column_spec(column=4, width="2cm") %>%
    kable_styling(latex_options="hold_position") %>%
    add_header_above(c(" "=1, "Adjusted effect"=2))
            
```

```{r bnfchapter1, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter1,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter1",
             col.names=c(clean.header(colnames(table.bnfchapter1)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 1 in those not resident in a care home and without without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```


```{r bnfchapter2, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter2,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter2",
             col.names=c(clean.header(colnames(table.bnfchapter2)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 2 in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
```

```{r bnfchapter4, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter4,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter4",
             col.names=c(clean.header(colnames(table.bnfchapter4)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 4 in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    group_rows("", 1, nrow(table.bnfchapter4)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()

```

```{r bnfchapter9, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.bnfchapter9,
             escape=FALSE, 
             booktabs=TRUE,
             label="bnfchapter9",
             col.names=c(clean.header(colnames(table.bnfchapter9)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align=rep("c", 6), 
             caption="Associations of severe disease with prescribed drugs in BNF chapter 9 in those without listed conditions") %>%
    column_spec(column=1, width="4cm") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    group_rows("", 1, nrow(table.bnfchapter9)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()

```

```{r protonpump, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.protonpump,
             escape=FALSE,
             label="protonpump.adjusted", 
             booktabs=TRUE,
             col.names=c(clean.header(colnames(table.protonpump)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe disease with prescription of a proton pump inhibitor by age group, with and without adjustment for hospital diagnosis of any upper GI tract disorder, and prescription of NSAIDs,  antiplatelet agent, and nonopioid analgesics, excluding care home residents") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()

```

```{r analgesic, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.nonopioid.analgesic,
             escape=FALSE,
             label="analgesic.adjusted", 
             booktabs=TRUE,
             col.names=c(clean.header(colnames(table.nonopioid.analgesic)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe disease with prescription of a nonopioid analgesic by age group, with and without adjustment for hospital diagnosis of any neoplasm, and prescription of  NSAIDs, antiplatelet agents and proton pump inhibitors, excluding care home residents") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()

```
