---
title: "Short report: efficacy of vaccination in those eligible for shielding in Scotland"
author:
  - name: Paul M McKeigue\textsuperscript{\getAff{Usher}}
    affiliation: HPS
  - name: David A McAllister\textsuperscript{\getAff{Glasgow}}  
    affiliation: HPS 
  - name: Jen Bishop
    affiliation: HPS
  - name: Sharon Hutchinson\textsuperscript{\getAff{GCU}} 
    affiliation: HPS
  - name: Chris Robertson\textsuperscript{\getAff{Strathclyde}}   
    affiliation: HPS
  - name: Nazir Lone
    affiliation: Usher
  - name: Jim McMenamin 
    affiliation: HPS
  - name: David Goldberg
    affiliation: HPS
  - name: Helen M Colhoun\textsuperscript{\getAff{IGMM}}
    corresponding: yes
    email: helen.colhoun@igmm.ed.ac.uk
    affiliation: HPS
address: 
- code: Usher
  address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care  
- code: HPS
  address: Public Health Scotland, Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
- code: Glasgow
  address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
- code: GCU
  address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
- code: Strathclyde
  address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
- code: IGMM
  address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - AXA Chair in Medical Informatics and Epidemiology

header-includes:
  \usepackage{newunicodechar}
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{geometry}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{longtable}
  \usepackage{lscape}
  \usepackage{pdfpages}
output: 
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    keep_tex: true
    fig_caption: yes
    md_extensions: -autolink_bare_uris
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
bibliography: ./covid.bib
csl: ./plos.csl
always_allow_html: true
urlcolor: blue
linenumbers: false
linkcolor: cyan
---


```{r vax, echo=FALSE, warning=FALSE, message=FALSE}

rollsum.datewin <- function(dt, k, datevar) {
    ## returns a table of rolling sums of width k centred on date
    ## dt is a dataset of totals (N) by date, which may have no rows for some dates in the
    ## range of interest
    ## add rows for missing dates by a left join of all.dates with dt
    all.dates <- data.table(date=seq(from=min(dt[[datevar]]),
                                     to=max(dt[[datevar]]),
                                     by=1))
    setkey(all.dates, date)
    # setkeyv(dt, datevar) can't set physical key here
    all.dates <- dt[all.dates]

    return(data.table(date=all.dates[[datevar]],
                      rsum=zoo::rollsum(all.dates[, N], k=k, fill=0, align="center")))
}

format.estcipv <- function(x.ci, x.pval) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    x <- gsub("times", "\\\\times", x)
    return(x)
}

format.estci <- function(x.ci) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    #x <- gsub("times", "\\\\times", x)
    return(x)
}

options(knitr.kable.NA = '.')
options(knitr.table.format = "latex") # adds tab: prefix to labels
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)

## vaccination analysis

datadir <- "data/2021-03-16/"
load(file=paste0(datadir, "cc.all.RData"))

shield.bydate <- cc.all[CASE==0 & SPECIMENDATE > as.Date("2020-12-01") & shield.any==TRUE,
                           .N,
                           by=c("SPECIMENDATE")]
setkey(shield.bydate, SPECIMENDATE)
shield.bydate <- rollsum.datewin(shield.bydate, 7, "SPECIMENDATE")
setnames(shield.bydate, "rsum", "all")

if(FALSE) {
    shield.bydate <- cc.all[CASE==0 & SPECIMENDATE > as.Date("2020-12-01") & shield.any==TRUE,
                            .N,
                            by=c("SPECIMENDATE", "vax14.dose")]
    setkey(shield.bydate, SPECIMENDATE)
    vax0.bydate <- rollsum.datewin(shield.bydate[vax14.dose==0], 7, "SPECIMENDATE")
    vax1.bydate <- rollsum.datewin(shield.bydate[vax14.dose==1], 7, "SPECIMENDATE")
    vax2.bydate <- rollsum.datewin(shield.bydate[vax14.dose==2], 7, "SPECIMENDATE")
    setkey(vax0.bydate, date)
    setkey(vax1.bydate, date)
    setkey(vax2.bydate, date)
    setnames(vax0.bydate, "rsum", "rsum0")
    setnames(vax1.bydate, "rsum", "rsum1")
    setnames(vax2.bydate, "rsum", "rsum2")
    shield.bydate <- vax2.bydate[vax1.bydate]
    shield.bydate <- shield.bydate[vax0.bydate]
}

shieldvax.bydate <- cc.all[CASE==0 & SPECIMENDATE > as.Date("2020-12-01") &
                           shield.any==TRUE & vax14==1,
                           .N,
                           by=c("SPECIMENDATE")]
setkey(shieldvax.bydate, SPECIMENDATE)
shieldvax.bydate <- rollsum.datewin(shieldvax.bydate, 7, "SPECIMENDATE")
setnames(shieldvax.bydate, "rsum", "vax")
setkey(shield.bydate, date)
setkey(shieldvax.bydate, date)       
shield.bydate <- shieldvax.bydate[shield.bydate]
shield.bydate[, prop.vax := vax / all]

p.shieldvax <-
    ggplot(data=shield.bydate, aes(x=date, y=prop.vax)) +
    geom_line() +
    scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), limits=c(0, 1),
                       expand=c(0, 0)) + 
    scale_x_date(breaks = seq.Date(from = as.Date("2020-12-15"),
                                   to =  as.Date("2021-03-02"), by = "week"),
                 labels=gsub("^0", "", 
                             format.Date(seq.Date(from = as.Date("2020-12-15"),
                                                  to =  as.Date("2021-03-02"),
                                                  by = "week"),
                                         "%d %b")
                             ),
                 limits=c(as.Date("2020-12-15"), as.Date("2021-03-02"))) +
        labs(x=paste0("Presentation date: mid-point of 7-day window"),
         y="Proportion vaccinated among controls eligible for shielding")

keep.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "prob.hcai", "care.home", "hh.over18", "listedgr3", "shield.any", "shield.group",
               listed.conditions, "shield.group", "vax14", "vax14.dose", "stratum")
cc.vax <- na.omit(cc.severe[SPECIMENDATE > as.Date("2020-12-01"), ..keep.vars])

keep.hosp.vars <- c("SPECIMENDATE", "CASE", "discharge15to56", "care.home", "shieldelig.group", 
                    "vax14", "stratum")
cc.hosp.vax <- na.omit(cc.all[(casegroup=="A" | casegroup=="B") &
                              SPECIMENDATE > as.Date("2020-12-01"), ..keep.hosp.vars])
rm(cc.all)

mh.strata <- cc.vax[care.home=="Independent" & shield.any==TRUE, 
                    .(SPECIMENDATE, stratum, CASE, vax14)]
mh.strata <- mh.strata[, c("a", "b", "c", "d") :=
                           list(CASE * as.integer(vax14),
                                (1 - CASE) * as.integer(vax14),
                                CASE * (1 - as.integer(vax14)),
                                (1 - CASE) * (1 - as.integer(vax14)))]
mh.strata <- mh.strata[, lapply(.SD, sum),
                       by=c("SPECIMENDATE", "stratum"), 
                       .SDcols=c("a", "b", "c", "d")]
mh.strata[, N := a + b + c + d]
mh.strata[, a.exp := (a + b) * (a + c) / N]
mh.strata[, c.exp := (a + c) * (c + d) / N]
mh.strata[, var.exp := (a + c) * (b + d) * (a + b) * (c + d) / (N^2 * (N - 1))]
mh.strata[is.nan(var.exp), var.exp := 0]
mh.strata[, week := paste(lubridate::year(SPECIMENDATE),
                          sprintf("%02d", lubridate::week(SPECIMENDATE)), sep=":")]

mh.dates <- mh.strata[, lapply(.SD, sum), by=week,
                      .SDcols=c("a", "a.exp", "c", "c.exp", "var.exp")]
setkey(mh.dates, week) 
mh.all <- data.table(week="All", mh.dates[, lapply(.SD, sum),
                                          .SDcols=c("a", "a.exp", "c", "c.exp", "var.exp")])
mh.dates <- rbind(mh.dates, mh.all)
colnames(mh.dates) <- c("Year:week", "Observed", "Expected", "Observed", "Expected", "Var(expected)")

table.dose <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + listedgr3 / vax14.dose + strata(stratum)))$coefficients

table.riskgroups <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + listedgr3 / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.riskgroups)
table.riskgroups <- as.data.table(table.riskgroups)
table.riskgroups[, rateratio := or.ci(coef, `se(coef)`)]
table.riskgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.riskgroups <- table.riskgroups[, .(rateratio, pvalue)]
table.riskgroups <- data.table(effect=effect, # freqs.riskgroups,
                               table.riskgroups)
table.riskgroups[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.riskgroups[, effect := gsub("^listedgr3", "", effect)]
table.riskgroups[, effect := gsub("TRUE$", "", effect)]
table.riskgroups[, effect := gsub(":vax14$", "", effect)]
table.riskgroups[, effect := replace.names(effect)]

table.adjhosp <- summary(clogit(data=cc.vax, 
                               formula = CASE ~ care.home + listedgr3 / vax14 +
                                   prob.hcai + discharge15to56 + strata(stratum)))$coefficients

table.adjhosp <- as.data.table(table.adjhosp)
table.adjhosp[, rateratio := or.ci(coef, `se(coef)`)]
table.adjhosp[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.adjhosp <- table.adjhosp[, .(rateratio, pvalue)]

table.shieldgroups <- summary(clogit(data=cc.vax, formula = CASE ~ care.home + shield.group / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.shieldgroups)
table.shieldgroups <- as.data.table(table.shieldgroups)
table.shieldgroups[, rateratio := or.ci(coef, `se(coef)`)]
table.shieldgroups[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.shieldgroups <- table.shieldgroups[, .(rateratio, pvalue)]
table.shieldgroups <- data.table(effect=effect, # freqs.shieldgroups,
                               table.shieldgroups)
table.shieldgroups[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.shieldgroups[, effect := gsub("^shield.group", "", effect)]
table.shieldgroups[, effect := gsub("TRUE$", "", effect)]
table.shieldgroups[, effect := gsub(":vax14$", "", effect)]
table.shieldgroups[, effect := replace.names(effect)]
table.shieldgroups <- as.data.frame(table.shieldgroups) 
                   
table.hosp <- summary(clogit(data=cc.hosp.vax, formula = CASE ~ care.home + shieldelig.group / vax14 + strata(stratum)))$coefficients
effect <- rownames(table.hosp)
table.hosp <- as.data.table(table.hosp)
table.hosp[, rateratio := or.ci(coef, `se(coef)`)]
table.hosp[, pvalue := format.pvalue(z, `Pr(>|z|)`)]
table.hosp <- table.hosp[, .(rateratio, pvalue)]
table.hosp <- data.table(effect=effect, # freqs.hosp,
                               table.hosp)
table.hosp[, effect := gsub("care.homeCare/nursing home", "Care home residence", effect)]
table.hosp[, effect := gsub("^shieldelig.group", "", effect)]
table.hosp[, effect := gsub("TRUE$", "", effect)]
table.hosp[, effect := gsub(":vax14$", "", effect)]
table.hosp[, effect := replace.names(effect)]
table.hosp <- as.data.frame(table.hosp) 

freqs.riskgroups <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "listedgr3"), data=cc.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "listedgr3"), data=cc.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]


oddsratio.table <- function(freqs.table) {
    freqs.table <- as.matrix(freqs.table)
    or <- numeric(nrow(freqs.table))
    for(i in 1:nrow(freqs.table)) {
        a <- as.numeric(gsub(" .+", "", freqs.table[i, ]))
        or[i] <- a[1] * a[4] / (a[2] * a[3])
    }
    return(round(or, 2))
}

freqs.shieldgroups <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "shield.group"), data=cc.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]
rownames(freqs.shieldgroups) <- table.shieldgroups$effect[9:16]
freqs.shieldgroups$"Crude odds ratio" <- oddsratio.table(freqs.shieldgroups)

freqs.hosp <- as.data.frame.matrix(cbind(
    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax,
                        drop.reflevel=FALSE),
    univariate.tabulate(varnames=c("care.home", "shieldelig.group"), data=cc.hosp.vax[vax14==1],
                        drop.reflevel=FALSE)))[-1, ]
rownames(freqs.hosp) <- table.hosp$effect[8:14]
freqs.hosp$"Crude odds ratio" <- oddsratio.table(freqs.hosp)


if(FALSE) {
cc.vax[`dm.typeOther/unknown type` == 1, `dm.typeType 2 diabetes` := 1]
cc.vax[, `dm.typeOther/unknown type` := NULL]

controls.vax <- cc.kept[CASE==0 & care.home=="Independent", .(AGE, COVID.age, vax_dose_1)]
controls.vax[, vaxnow := !is.na(vax_dose_1)]
controls.vax[, vaxnow := car::recode(vaxnow,
                                   recodes="FALSE='Not vaccinated';
                                            TRUE='At least one dose'",
                                   as.factor=TRUE,
                                   levels=c("Not vaccinated", "At least one dose"))]

theme_set(theme_gray(base_size = 16))
p.covidage <- ggplot(data=controls.vax, aes(x=AGE, y=COVID.age, color=vaxnow)) +
    geom_point(size=0.5, position="jitter") +
    scale_x_continuous(breaks=seq(20, 80, by=10), limits=c(15, 90)) + 
    scale_y_continuous(breaks=seq(20, 80, by=10), limits=c(15, 90)) +
    theme(legend.title = element_blank()) +
    theme(legend.position=c(0.15, 0.85)) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    labs(title=paste("COVID age versus calendar age by vaccination status on",
                        format.Date(max(vacc$vax_dose_1), '%d %b %Y')),
         caption="Controls matched to severe cases, not resident in care homes. Limits of COVID age are 16 and 90 years",
         x="Calendar age (years)",
         y="COVID age (years)")
p.covidage

png("COVIDage_vaxstatus.png")
p.covidage
dev.off()
rm(p.covidage)

vax.firstdoses <- vacc[, .N, by=c("weekdose1", "vax_type_1")]
ggplot(data=vax.firstdoses, aes(x=weekdose1, y=N, color=vax_type_1)) +
    geom_line() 

}

```

# Background
In Scotland to date COVD-19 vaccination has been with either the ChAdOx1 nCoV-19 (Astra-Zeneca) or BNT162b2 mRNA (Pfizer-BioNtech) vaccines.  Both have been shown to be efficacious in clinical trials.  Real-world observational analysis in a primary care-based study in Scotland up to 15 Feb 15 2021 found maximal efficacy of 70% for ChAdOx1 nCoV-19 and 85% for BNT162b2 mRNA at 28 days post-vaccination [@vasileiou_effectiveness_2021].  However the efficacy of COVID-19 vaccination among those designated as clinically extremely vulnerable is still uncertain; an official [letter](https://www.gov.scot/publications/coronavirus-covid-19-shielding-letters/) sent on 23 March 2021 to those on this list stated that "we do not yet have evidence of exactly how effective [the vaccine] is for the people on the shielding list."   This is relevant to whether individuals in this category should be advised to remain in shielding as restrictions on the general population are lifted.  This study examines the efficacy of vaccination in reducing the risk of severe COVID-19 among those designated as clinically extremely vulnerable and thus eligible for shielding in Scotland.

# Methods 
This analysis is based on the REACT-SCOT matched case-control study, established at the beginning of the epidemic to explore risk factors for severe COVID-19 in the population of Scotland.  The design has been described in detail previously [@mckeigue_rapid_2020]. In brief, for every incident case of COVID-19 in the population ten controls matched to one-year age, sex and primary care practice were selected using the Community Health Index database.  COVID-19 cases are those with a positive nucleic acid test, or a hospital admission of death with COVID-19 codes. The dataset is refreshed regularly and is linked to the vaccination database and to the regularly updated dataset of all individuals deemed eligible for the shielding programme.  This analysis is based on all cases presenting between 1 December 1st 2020 to 16 March 2021.

We used this matched case-control design to take advantage of extensive data linkages already established. As the design uses incidence density sampling, it controls not only for the matching factors of age, sex and general practice but also for calendar time which would otherwise confound the association between vaccination and disease. 

As previously, to minimise ascertainment bias we focused the analysis on  severe COVID-19,  defined as entry to critical care within 28 days of presentation or  fatal outcome (any death within 28 days of a positive test or any death for which COVID-19 was coded as underlying cause).   As described previously, cases and controls were grouped into one of three categories: no risk condition, risk condition designated previously by Public Health Scotland as moderate risk, eligible for shielding.  For further analyses, the shielding category was subdivided as described previously into six categories: solid organ transplant, specific cancers, severe respiratory conditions, rare conditions, on immunosuppressants, additional conditions.  

Vaccination status in cases was coded as having had at least one dose of any vaccine at least 14 days before presentation date. Ths assigns cases (and their matched controls) who were vaccinated less than 14 days prior to the case date to the unexposed category.  The numbers of cases and controls were too few too allow further subdivision of time since vaccination.  

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model fitted to the full dataset, specifying effects $\beta_{R2}, \dots, \beta_{RJ}$ for the log rate ratio associated with risk categories 2 to $J$ ($\beta_{R1} = 0$ for the reference category $J=1$), and "nested" effects $\beta_{V1}, \dots, \beta_{VJ}$ for the log rate ratio associated with vaccination in each of the $J$ risk categories.  Because severe COVID-19 is strongly associated with care home residence and care home residents were a priority group for vaccination, the model includes the log rate ratio $\beta_H$ associated with care home residence. For the $i$th individual in a matched set whose care home residence status is $H_i$, is in risk category $j$ and has vaccination status $V_i$, the linear predictor $\Theta_i$ is $\beta_H H_i + \beta_{Rj} + \beta_{Vj} V_{i}$.  The conditional likelihood for a matched set where individual $c$ is the case is $\exp{\left( \Theta_c \right) } / \sum_i{\exp{\left( \Theta_i \right) }}$. The efficacy of vaccination in the $j$th risk category is 1 minus the rate ratio $\exp{\left(\beta_{Vj} \right)}$.  

As recent hospital admissions or being in hospital before presentation date might confound the association of vaccination with severe COVID-19, the analysis was repeated with adjustment for any hospital discharge from 2 to 8 weeks before presentation date and for being an inpatient for at least 8 days before presentation date.  Because As another sensitivity analysis and to allow inference on vaccine efficacy in less frequent risk groups, the analysis was repeated with the case definition extended to include all cases that were hospitalized or fatal.  For this analysis the coding of moderate risk conditions was not available, so that those with no risk condition and those with moderate risk conditions are instead grouped together as "ineligible for shielding".  

# Results
Figure \ref{fig:plotvax} shows the  proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date.  This proportion in  controls  will be a close estimate of the  overall  vaccination rates in the total shielding population.  By 9 February 2021 about half the shielded population were vaccinated, rising to over 85% by early March.  

```{r plotvax, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=4, fig.asp=0.8, fig.cap="\\label{fig:plotvax}Proportion vaccinated at least 14 days before among those eligible for shielding in the control group, by presentation date"}

theme_set(theme_gray(base_size = 8))
p.shieldvax

```

Supplementary Table \ref{tab:freqsshieldgroups} shows the distribution of risk groups among  all severe cases and controls and then vaccinated cases and controls. Supplementary Table \ref{tab:freqshosp} shows the same breakdown for hospitalised cases and controls.   These   tables show that the numbers of cases in some shielding subgroups, especially solid organ transplant recipients, are too small for rate ratios to be estimated reliability. Supplementary Table \ref{tab:mhdates} shows the observed and expected numbers of severe cases among vaccinated cases and unvaccinated individuals who eligible for shielding and not resident in care homes, by week of presentation.  This shows that after week 6 of 2021 the number of severe cases was greatere in vaccinated than in unvaccinated individuals.  This is to be expected as by this time more than 70% of those eligible for shielding had been vaccinated; it does not indicate that vaccination was ineffective. 

The rate ratios for severe COVID-19 associated with vaccination were `r format.estci(table.riskgroups$rateratio[4])` in those with no risk condition, `r format.estci(table.riskgroups$rateratio[5])` in those with moderate risk conditions and `r format.estci(table.riskgroups$rateratio[6])` in those eligible for shielding.  With adjustment for having been in hospital for more than 7 days before presentation and for any hospital discharge from 15 to 56 days before presentation, the corresponding rate ratios were `r format.estci(table.adjhosp$rateratio[6])`, `r format.estci(table.adjhosp$rateratio[7])` and `r format.estci(table.adjhosp$rateratio[8])`.  
Table \ref{tab:shieldgroups} gives a breakdown by shielding group.  For the categories labelled as specific cancers, severe respiratory disease and additional conditions the numbers are large enough for the effect of vaccination on severe disease to be estimated reliably.  The point estimates for the rate ratio for all risk groups were  lower than for those without risk conditions i.e. vaccine efficacy was at least as high,  with the exception of  solid organ transplant recipients,  those  with rare diseases and  those on immunosuppressants.  For these groups the numbers of cases and controls were small (Supplementary Table \ref{tab:freqsshieldgroups}) and confidence intervals were wide. 


```{r shieldgroups, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.shieldgroups[9:16, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldgroups",
             row.names=FALSE,
             align=c("l", "r", "l"),
             caption="Rate ratios for severe COVID-19 associated with vaccination, within each risk group",
             col.names=c("Effect", 
                         "Rate ratio (95\\% CI)", "\\ensuremath{p}-value")) %>%
    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
                         "Vaccine effects nested within each level of risk group",
                         "Care home residence included as covariate", 
                         "Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(column=1, width="6cm") %>%
    kable_styling(latex_options=c("HOLD_position")) 


```

Table \ref{tab:shieldelig} shows the analysis extended to include all hospitalized cases, for which the numbers of cases are larger.   This shows significant efficacy in all groups other than solid organ transplant recipients but in this group the numbers of cases and controls were small even with the broader case definition (Supplementary Table \ref{tab:freqshosp}). 

```{r hosp, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.hosp[8:14, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="shieldelig",
             row.names=FALSE,
             align=c("l", "r", "l"),
             caption="Rate ratios for hospitalized or fatal COVID-19 associated with vaccination within each risk group",
             col.names=c("Effect", "Rate ratio (95\\% CI)", 
                         "\\ensuremath{p}-value")) %>%
    add_footnote(label=c("Conditional logistic regression model matched on age, sex, general practice and presentation date",
                         "Care home residence included as covariate", 
                         "Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    kable_styling(latex_options=c("HOLD_position")) 

```

# Conclusions
Efficacy of a single vaccination dose in protecting against severe COVID-19 is as high or higher in those eligible for shielding as in those without risk conditions.  For solid organ transplant recipients -- the group at highest risk among those eligible for shielding -- the number of cases is still too small for the efficacy of vaccination to be estimated.  

The estimates of the rate ratio associated with vaccination or the vaccine efficacy here cannot be directly compared with the previous report for the total population of Scotland as they are summary estimates across all time periods from 14 days post-vaccination, rather than over 7-day intervals.  Nonetheless the estimate is close to that reported at 14 days in the  previous report.

Since we have previously shown that  the shielding group overall have approximately 6 fold higher risk of severe COVID-19 than those of the same age and sex without risk conditions [@mckeigue_relation_2021-1] they would need a vaccine efficacy of  82% (i.e. a rate ratio of 0.18)  with a single dose to achieve a lowering of risk to the level of unvaccinated persons  without risk conditions.   Thus whilst those in shielding  can be reassured that the vaccine is showing efficacy they should also be advised that they remain at increased risk  after a single dose (about twofold compared to those of the same age and sex without risk conditions given the vaccine-associated rate ratio of about 0.3). Thus they should remain vigilant  until they have received their second dose and background transmission rates are low. Also a  larger dataset than the  Scottish population is needed to estimate efficacy in solid organ transplant recipients where we have not been able to demonstrate efficacy. 

The vaccine Green Book recommends that the second dose of all  COVID-19 vaccines should be routinely scheduled between four and 12 weeks after the first dose [@public_health_england_covid-19_2021].  Although earlier rather than later second dosing within that interval might more quickly reduce the residual risk among those in shielding,   this is not necessarily the optimal strategy as the JCVI have noted that "With most vaccines an extended interval between the prime and booster doses leads to a better immune response to the booster dose." [@joint_committee_on_vaccination_and_immunisation_optimising_2021]

# References
<div id="refs"></div>

\beginsupplement
\onehalfspacing

# Supplementary Material

```{r freqsshieldgroups, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(freqs.shieldgroups, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqsshieldgroups",
             row.names=TRUE,
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of severe COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none") %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position")) 

```

```{r freqshosp, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(freqs.hosp, 
             escape=FALSE, 
             booktabs=TRUE,
			 label="freqshosp",
             row.names=TRUE,
             align=c("l", "r", "r", "r", "r"),
             caption="Numbers of controls and cases of hospitalized or fatal COVID-19 by risk group and vaccination status") %>%
    add_header_above(c(" "=1, "All"=2, "Vaccinated"=2)) %>% 
    add_footnote(label=c("Presentation dates from 1 December 2020 to 16 March 2021",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    column_spec(2, width="2cm") %>%
    column_spec(3, width="2cm") %>%
    column_spec(4, width="2cm") %>%
    column_spec(5, width="2cm") %>%
    column_spec(6, width="1cm") %>%
    kable_styling(latex_options=c("hold_position")) 


```

Vaccine efficacy cannot be estimated from the crude odds ratios in Tables \ref{tab:freqsshieldgroups} and  \ref{tab:freqshosp} because these totals do not take account of the matched case-control design.   The matched design makes it possible to control for age, sex, general practice and crucially for calendar date where the proportion vaccinated and the rates of severe disease were changing day by day during the study period.  The vaccine efficacy is estimated over matched sets, conditional on the marginal totals (cases vs controls, vaccinated versus unvaccinated) in each set. This section is intended to provide a more intuitive explanation of this.  Table \ref{tab:mhdates} shows how information for a statistical test (the Mantel-Haenszel test) of  vaccine efficacy accumulated over matched sets (strata) of cases and controls during the study period.  To keep the working simple, this table is restricted to those not resident in care homes so there is no covariate adjustment.  

Each matched set can be represented as a 2 \ensuremath{\times} 2 table, in which the numbers in each cell are traditionally labelled $a, b, c, d$: 
 
 | | Case | Control | All |
 | --- | --- | --- | --- |
 Vaccinated | $a$ | $b$ | $V_1$ |
 Unvaccinated | $c$ | $d$ | $V_0$ |
 All | $C_1$ | $C_0$ | $N$ |

The marginal totals of cases and controls are $C_1, C_0$ and of vaccinated and unvaccinated are $V_1, V_2$.  Conditional on these marginal totals the expected number of vaccinated cases under the null hypothesis of no association is $C_1 V_1 / N$ and the variance of the expected is $C_1 C_0 V_1 V_0 / \left[ N^2 \left(N - 1 \right) \right]$.  The observed numbers, the expected and the variance are summed over matched sets to calculate a test statistic as (observed - expected) divided by the square root of the variance.  


```{r mh.dates, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(mh.dates[1:17, ], 
             escape=FALSE, 
             booktabs=TRUE,
			 label="mhdates",
             row.names=FALSE,
             align=c("l", "r", "r", "r", "r"),
             digits=1,
             caption="Vaccinated and unvaccinated severe cases among those eligible for shielding and not resident in care homes: observed and expected numbers by week") %>%
    add_header_above(c(" "=1, "Vaccinated cases"=2, "Unvaccinated cases"=2)) %>% 
    add_footnote(label=c("Expected numbers (under the null) are calculated in each stratum",
                         "Vaccine status coded as 1 if at least one dose at least 14 days before, 0 otherwise"), 
                 notation="none")  %>%
    kable_styling(latex_options=c("hold_position")) 

```
